Pre- and Post-Debate Democratic Primary Data: Twitter, Google Trends, and Polls

### 2019-12-10 Paper/paper-draft-Padgett-Thiam.Rmd
Update paper-draft-Padgett-Thiam.Rmd

---
title: "Pre- and Post-Debate Democratic Primary Data: Twitter, Google Trends, and Polls"
subtitle: "Fundamentals of Computing and Data Display, Fall 2019"
author: "Zoe Padgett and Fatou Thiam"
date: "`r Sys.Date()`"
fontsize: 12pt
text-align: left
output:
  pdf_document:
    toc: yes
    df_print: kable
references:
- id: GayoAvello2013
  title: A Meta-Analysis of State-of-the-Art Electoral Prediction From Twitter Data
  author:
  - family: Gayo-Avello
    given: D.
  container-title: Organization Studies
  volume: 31
  issue: 6
  page: 211–248
  type: article-journal
  issued:
    year: 2013
- id: Beauchamp2017
  title: Predicting and Interpolating State-Level Polls Using Twitter Textual Data
  author:
  - family: Beauchamp
    given: N.
  container-title: American Journal of Political Science 
  volume: 61
  issue: 2
  type: article-journal
  issued:
    year: 2017
- id: Gaurav2013
  title: Leveraging Candidate Popularity on Twitter to Predict Election Outcome
  author:
  - family: Gaurav
    given: M.
  - family: Srivastava
    given: A.
  - family: Kumar
    given: A.
  - family: Miller
    given: S.
  container-title: Proceedings of the 7th Workshop on Social Network Mining and Analysis
  type: article-journal
  issued:
    year: 2013
- id: Kassraie2017
  title: Election Vote Share Prediction using a Sentiment-based Fusion of Twitter Data with Google Trends and Online Polls
  author:
  - family: Kassraie
    given: P.
  - family: Modirshanechi
    given: A.
  - family: Aghajan
    given: H.
  container-title: Proceedings of the 6th International Conference on Data Science, Technology and Applications
  type: article-journal
  issued:
    year: 2017
- id: Salunkhe2017
  title: Twitter Based Election Prediction and Analysis
  author:
  - family: Salunkhe
    given: P.
  - family: Deshmukh
    given: S.
  container-title: International Research Journal of Engineering and Technology
  volume: 4
  issue: 10
  type: article-journal
  issued:
    year: 2017
- id: Gtrendscite   
  title: "gtrendsR: Perform and Display Google Trends Queries"
  author:
  - family: Massicotte
    given: P.
  - family: Eddelbuettel
    given: D.
  issued:
    year: 2019
  note: R package version 1.4.4.9000
  url: https://github.com/PMassicotte/gtrendsR
- id: shinycite
  title: "shiny: Web Application Framework for R"
  author:
  - family: Chang
    given: W.
  - family: Cheng 
    given: J.
  - family: Allaire
    given: J.
  - family: Xie
    given: Y. 
  - family: McPherson
    given: J.
  issued:
    year: 2019
  note: R package version 1.4.0
  url: https://CRAN.R-project.org/package=shiny
- id: mapcite
  title: "ggmap: Spatial Visualization with ggplot2"
  author:
  - family: Kahle
    given: D.
  - family: Wickham
    given: H.
  container-title: The R Journal
  volume: 5
  issue: 1
  page: 144-161
  type: article-journal
  issued:
    year: 2013
  url: https://journal.r-project.org/archive/2013-1/kahle-wickham.pdf
---

@@

Using the rtweet package and the search_tweet function, we were able to gather sufficient tweets for our project and classified it as pre or post debate. The initial dataset contains a little over 1.3 million tweets which was rich enough to proceed with analysis. After removing irrelevant tweets such as tweets containing "Ukraine", we were left with 1.2 million tweets which we then divided in two groups :pre and post debate. The pre-debate dataset has 226,799 tweets and 998,780 tweets for the post-debate dataset. As expected, there was a lot more tweets after the debate than before. Twitter has a location field since we wanted to focus our anylysis on certain states in the U.S., however the location column is an open text and not regulated. We selected severals states: New York, Virginia, New Hampshire, Nevada, Illinois, Iowa, Ohio, California and DC to focus our analysis on.

@@

Using the search_tweets function in the rtweets package, we will be quering tweets with the keywords specified below. We limited our tweets pool to November 20th and November 21st. The reasonning behind this is to get a feel post and pre debate.

@@

Even after we specify the keywords, we know that we would still get irrelevants tweets mainly with the Ukraine issue at the time of pulling. We removed all tweets that had the word "Ukraine" in it.

```{r, eval=FALSE}
myvars <- c("text", "location", "created_at")
df1 <- dem[myvars]
df2 <- dem2[myvars]
```

@@

Next, we seperate the tweets by candidates in order to do the analysis at the candidate level. It's important to note that one tweets can be addressed to multiple candidates. In that case, the tweet will be found in the each of those candidate dataset. The code below exemplifies this step of the cleaning process.

@@

Here we reformat the created_at column as a date & time variable in order to seperate the tweets in two groups : post and pre debate.
```{r, eval=FALSE}
tweets <- tweets %>%
  mutate( date= as.POSIXct(created_at, tryFormats = c("%Y-%m-%d %H:%M:%OS")))

@@

All the tweets received before 9 p.m. on November 20th are accounted for in the pre-debate dataset and all tweets from 11 p.m. on November 20th to the next day are in the post-debate dataset. Note that tweets during the debate( 9 - 11 p.m are ignored)
```{r, eval=FALSE, tidy=TRUE}
pre_debate <- tweets %>%
              filter(date(date) == "2019-11-20" & hour(date) < 21 )

@@

#### Sentiment Analysis
Sentiment analysis of the tweets will performed for only 5 candidates.

First we need to prep the tweets by removing all non-words such as emojis and use the sentiment analysis package for analysis. Althought the sentiment analysis packae uses 5 dictionnairies, we will only look at the results from the GI dictionnary. For the pre-debate tweets sentimeent we used all the tweets from the dataset; however for the post-debate analysis, we sampled from the dataset due to volume and processing error. The code below exemplifies this process.   
```{r, eval=FALSE}
bs <- pre_debate %>%
      filter (BS == "TRUE")

@@

@@

```
Figure 1 shows the number of tweets that mentioned each candidate before the debate. We noticed that Sanders, Biden and Buttigieg are in the lead with a little over 25,000 mentions.

@@

Figure 2 shows the post- debate tweets mentions by candidate. As expected lots more tweets mentions occured after the debate, 5 times more than pre- debate. Here we noticed that the ranking change with Harris and Yang in the lead along with Sanders.

@@

```
Figure 3 .....
## Tweets location post-debate 
```{r, fig.cap="Post-debate Mentions in Tweets, by Candidate & location"}

@@

@@

This section summarizes the results and may briefly outline advantages and limitations of the work presented.

@@

 Kearney, M. W. (2018). rtweet: Collecting Twitter Data. R package version 0.6.7 
 Retrieved from https://cran.r-project.org/package=rtweet

 Silge J, Robinson D (2016). “tidytext: Text Mining and Analysis Using Tidy Data Principles in R.” _JOSS_, *1*(3). 
 doi: 10.21105/joss.00037 (URL http://doi.org/10.21105/joss.00037), <URL:http://dx.doi.org/10.21105/joss.00037>.

 D. Kahle and H. Wickham. ggmap: Spatial Visualization with ggplot2.
 The R Journal, 5(1), 144-161. URL http://journal.r-project.org/archive/2013-1/kahle-wickham.pdf

 Garrett Grolemund, Hadley Wickham (2011). Dates and Times Made Easy with
 lubridate. Journal of Statistical Software, 40(3), 1-25. URL http://www.jstatsoft.org/v40/i03/.

 Benoit K, Watanabe K, Wang H, Nulty P, Obeng A, Müller S, Matsuo A (2018).  “quanteda: An R package for the quantitative
 analysis of textual data.” _Journal of Open Source Software_, *3*(30), 774. doi: 10.21105/joss.00774 
 (URL: http://doi.org/10.21105/joss.00774), <URL: https://quanteda.io>.

  Stefan Feuerriegel and Nicolas Proellochs (2019). SentimentAnalysis: Dictionary-Based Sentiment Analysis. 
  R package version 1.3-3. https://CRAN.R-project.org/package=SentimentAnalysis

  Barret Schloerke, Jason Crowley, Di Cook, Francois Briatte, Moritz Marbach, Edwin Thoen, 
  Amos Elberg and Joseph Larmarange (2018). GGally:  Extension to 'ggplot2'. R package version 1.4.0. 
  https://CRAN.R-project.org/package=GGally

   H. Wickham. ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag New York, 2016.

  Gergely Daróczi and Roman Tsegelskyi (2018). pander: An R 'Pandoc'  Writer. R package version 0.6.3.
  https://CRAN.R-project.org/package=pander

>>>>

---
title: "Pre- and Post-Debate Democratic Primary Data: Twitter, Google Trends, and Polls"
subtitle: "Fundamentals of Computing and Data Display, Fall 2019"
author: "Zoe Padgett and Fatou Thiam"
date: "`r Sys.Date()`"
fontsize: 12pt
text-align: left
output:
  pdf_document:
    toc: yes
    df_print: kable
references:
- id: GayoAvello2013
  title: A Meta-Analysis of State-of-the-Art Electoral Prediction From Twitter Data
  author:
  - family: Gayo-Avello
    given: D.
  container-title: Organization Studies
  volume: 31
  issue: 6
  page: 211–248
  type: article-journal
  issued:
    year: 2013
- id: Beauchamp2017
  title: Predicting and Interpolating State-Level Polls Using Twitter Textual Data
  author:
  - family: Beauchamp
    given: N.
  container-title: American Journal of Political Science 
  volume: 61
  issue: 2
  type: article-journal
  issued:
    year: 2017
- id: Gaurav2013
  title: Leveraging Candidate Popularity on Twitter to Predict Election Outcome
  author:
  - family: Gaurav
    given: M.
  - family: Srivastava
    given: A.
  - family: Kumar
    given: A.
  - family: Miller
    given: S.
  container-title: Proceedings of the 7th Workshop on Social Network Mining and Analysis
  type: article-journal
  issued:
    year: 2013
- id: Kassraie2017
  title: Election Vote Share Prediction using a Sentiment-based Fusion of Twitter Data with Google Trends and Online Polls
  author:
  - family: Kassraie
    given: P.
  - family: Modirshanechi
    given: A.
  - family: Aghajan
    given: H.
  container-title: Proceedings of the 6th International Conference on Data Science, Technology and Applications
  type: article-journal
  issued:
    year: 2017
- id: Salunkhe2017
  title: Twitter Based Election Prediction and Analysis
  author:
  - family: Salunkhe
    given: P.
  - family: Deshmukh
    given: S.
  container-title: International Research Journal of Engineering and Technology
  volume: 4
  issue: 10
  type: article-journal
  issued:
    year: 2017
- id: Gtrendscite   
  title: "gtrendsR: Perform and Display Google Trends Queries"
  author:
  - family: Massicotte
    given: P.
  - family: Eddelbuettel
    given: D.
  issued:
    year: 2019
  note: R package version 1.4.4.9000
  url: https://github.com/PMassicotte/gtrendsR
- id: shinycite
  title: "shiny: Web Application Framework for R"
  author:
  - family: Chang
    given: W.
  - family: Cheng 
    given: J.
  - family: Allaire
    given: J.
  - family: Xie
    given: Y. 
  - family: McPherson
    given: J.
  issued:
    year: 2019
  note: R package version 1.4.0
  url: https://CRAN.R-project.org/package=shiny
- id: mapcite
  title: "ggmap: Spatial Visualization with ggplot2"
  author:
  - family: Kahle
    given: D.
  - family: Wickham
    given: H.
  container-title: The R Journal
  volume: 5
  issue: 1
  page: 144-161
  type: article-journal
  issued:
    year: 2013
  url: https://journal.r-project.org/archive/2013-1/kahle-wickham.pdf
- id: Gtrendscite   
  title: "gtrendsR: Perform and Display Google Trends Queries"
  author:
  - family: Massicotte
    given: P.
  - family: Eddelbuettel
    given: D.
  issued:
    year: 2019
  note: R package version 1.4.4.9000
  url: https://github.com/PMassicotte/gtrendsR
- id: rtweet
  title: "rtweet: Collecting Twitter Data"
  author:
  - family: Kearney
    given: M.
  issued:
    year: 2019
  note: R package version 0.6.9
  url: https://cran.r-project.org/package=rtweet
- id: tidytext
  title: "tidytext: Text Mining and Analysis Using Tidy Data Principles in R"
  author:
  - family: Silge
    given: J.
  - family: Robinson
    given: D.
  doi: 10.21105/joss.00037
  url: http://dx.doi.org/10.21105/joss.00037
  issued:
    year: 2016
  publisher: The Open Journal
  note: R package version 0.6.9
  volume: 1
  number: 3
  journal: JOSS
- id: sentiment
  title: "SentimentAnalysis: Dictionary-Based Sentiment Analysis"
  author:
  - family: Feuerriegel
    given: S.
  - family: Proellochs
    given: N.
  doi: 10.21105/joss.00037
  url: https://CRAN.R-project.org/package=SentimentAnalysis
  issued:
    year: 2019
  note: R package version 1.3-3
- id: quanteda
  title: "quanteda: An R package for the quantitative analysis of textual data"
  author:
  - family: Benoit
    given: K.
  - family: Watanabe
    given: K.
  - family: Wang
    given: H.
  - family: Nulty
    given: P.
  - family: Obeng
    given: A.
  - family: Müller
    given: S.
  - family: Matsuo
    given: A.
  doi: 10.21105/joss.00774
  url: https://quanteda.io
  issued:
    year: 2018
  volume: 3
    number: 30
    pages: 774
- id: tidyverse
  title: "tidyverse: Easily Install and Load the 'Tidyverse'"
  author:
  - family: Wickman
    given: H.
  url: https://CRAN.R-project.org/package=tidyverse
  issued:
    year: 2017
  note: R package version 1.2.1

---

@@

Using the rtweet package and the search_tweet function, we were able to gather sufficient tweets for our project and seperated in two datasets; pre- and post- debate. The initial dataset contains a little over 1.3 million tweets which was rich enough to proceed with analysis. After removing irrelevant tweets such as tweets containing "Ukraine", we were left with over 1.2 million tweets. This was used to create the pre- and post- debate datasets. The pre-debate dataset has 226,799 tweets and 998,780 tweets for the post-debate dataset. As expected, there was a lot more tweets after the debate than before. We used the  Twitter location field to perform analysis at the state level.However, the location column is an open text and not regulated. We selected severals states such New York, Virginia, New Hampshire, Nevada, Illinois, Iowa, Ohio, California and DC to focus our analysis on.

@@

Using the search_tweet function in the rtweet package, we will be quering tweets with the specified keywords below. We limited our tweets pull to November 20th and November 21st. The reasonning behind this is to get a sense on how Twitter users feel pre- and post- debate.

@@

Even with specified keywords, irrelevants tweets were also made it to the initial dataset mainly with the Ukraine issue at the time of pulling. We removed all tweets that had the word "Ukraine" in order to focus completly on tweets regarding the debate.

```{r, eval=FALSE}
vars <- c("text", "location", "created_at")
df1 <- dem[vars]
df2 <- dem2[vars]
```

@@

Next, we seperate the tweets by candidates in order to do analysis at the candidate level. It's important to note that one tweets can be addressed to multiple candidates. In that case, the tweet will be found in the each of those candidates dataset. The code below exemplifies this step of the cleaning process.

@@

Here we reformat the created_at column as a date & time variable in order to seperate the tweets in two groups : pre- and post- debate.
```{r, eval=FALSE}
tweets <- tweets %>%
  mutate( date= as.POSIXct(created_at, tryFormats = c("%Y-%m-%d %H:%M:%OS")))

@@

All the tweets received before 9 p.m. on November 20th were accounted for in the pre-debate dataset and all tweets from 11 p.m. on November 20th to the next day are in the post-debate dataset. Note that tweets during the debate( 9 - 11 p.m. were removed)
```{r, eval=FALSE, tidy=TRUE}
pre_debate <- tweets %>%
              filter(date(date) == "2019-11-20" & hour(date) < 21 )

@@

#### Sentiment Analysis
Sentiment analysis of the tweets will not be performed for all candidates but only 5 : Harris, Sanders, Biden, Warren & Buttigieg.

First we need to prep the tweets by removing all non-words such as emojis and use the sentiment analysis package for analysis. Althought the sentiment analysis package uses 5 dictionnairies, we will only look at the results from the GI dictionnary. For the pre-debate tweets sentiment analysis, we used all the tweets from the dataset. However, for the post-debate sentiment analysis we sampled from the dataset due to volume and space issue. The code below exemplifies this process. The sample size for each candidate post-debate dataset is the same as the pre-debate number of tweets for consistency. For instance, the pre-debate dataset for Sanders had 31,603 tweets, the sample size for the post-debate dataset for Sanders was 31,603.     
```{r, eval=FALSE}
bs <- pre_debate %>%
      filter (BS == "TRUE")

@@

Table 1 shows the number of mentions by candidate pre- and post- debate and Table 2 shows similar results adding location as a factor. In both tables, we see that there were lot more tweets after the debate. This was anticipated with the debate being a headline the next day. This was also the reason why we decided to focus on pre- and post- debate analysis and not focus on the number of tweets. Sentiment analysis was performed using the GI dictionnaire in the sentiment analysis package [@sentiment].Each tweet received a positive, neutral and/ or negative score. For this project, we summed the positive score of all tweets and divided by the number of tweets in each dataset. This gave us the average tweet positivity by candidate shown in Table 3. Candidate tweet positivity pre- debate are very close with xx% for candx being the highest. We notice a slight increase for almost all candidates with now xx% forcandx in the lead.

@@

```
Figure 1 shows the number of tweets that mentioned each candidate before the debate. We noticed that Sanders, Biden and Buttigieg are in the lead with over 25,000 tweets mentions.

@@

Figure 2 shows the post- debate tweets mentions by candidate. As expected, lot more tweets mentions occured after the debate, about 5 times more than pre- debate. Here we noticed that the ranking change with Harris and Yang in the lead along with Sanders and tweets mentions close to 150,000.

@@

```
Figure 3 shows the number of tweets mentions by candidate by location pre- debate. Here we notice that Buttigieg, Sanders and Yang get the most tweets mentions and mainly coming from California and New York for all 3 candidates. These results are not surprising with California and New York population presence in Twitter.

## Tweets location post-debate 
```{r, fig.cap="Post-debate Mentions in Tweets, by Candidate & location"}

@@

```
Figure 4 is the post- debate tweets mentions by candidate by location. Just like in Figure 3 (pre-debate analysis) California and New York had the most users tweeting about the candidates.Harris, Yang and Sanders are mostly mentionned in tweets by users.

@@

It would be advantageous to survey methodologists to be able to use sources such as Twitter to replace survey data. However, we need to aknowledge the limitations preventing us from doing so. We all will agree that Twitter users do not adequately represent the population consequently introducing coverage error. Validity and measurement error just to name a few limitations we encountered with Twitter data. In lieu of replacing survey data, Twitter data can/ should be use as a supplemental data. It is important to note that we collected data for a short period of time. We believed that Twitter data can be much more useful if we had a longer timeframe. With that we would be able to perform more in depht analysis with time a key variable or even fit models. For instance, the next democratic debate is scheduled for December 19th, it would be intersting to see how the sentiment differ from the November 21st debate. On that point, candidates could use Twitter data for many other purposes such as percpetion in the Twitter world or change over time in sentiment but definitely not for election outcomes.

@@
